---
title: "lab07"
format: html
---

Import libraries

```{r}
library(tidyverse)
library(broom) 
library(cowplot)
```

Import biopsy database

```{r}
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")
```

Only keep numeric values and scale

```{r}
pca_fit <- biopsy %>% 
  select(where(is.numeric)) %>%
  prcomp(scale = TRUE) 
```

Visualization fun

```{r}
options(bitmapType = "cairo")
pca_fit %>%
  augment(biopsy) %>% 
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#FF00FF", benign = "#0000FF")
  ) +
  theme_half_open(12) + background_grid()
```

Plot 2:

```{r}
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit %>%
  tidy(matrix = "rotation") %>%
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#FF00FF"
  ) +
  xlim(-1.25, .5) + ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid(12)
```

Plot 3:

The goal is to plot the proportion of variance explained by each principal component.

```{r}
pca_fit %>%
  tidy(matrix = "eigenvalues") %>%
  ggplot(aes(PC, percent, fill = factor(PC == 1))) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(
    values = c("TRUE" = "#D55E00", "FALSE" = "#56B4E9"),
    guide = "none" 
  ) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    title = "Variance Explained by Principal Components",
    x = "Principal Component",
    y = "Variance Explained (in %)"
  ) +
  theme_minimal_hgrid(12)


pca_fit <- biopsy %>% 
  select(where(is.numeric)) %>%
  prcomp(scale = TRUE) 
```

We see that PC1 (displayed in orange) explains roughly 65% of the variance, which is why it separates benign vs malignant samples well.

\
The other PCs (displayed in blue) explain less and less indicating that the contribution is lower for other PCs.
